Foundational claims of group polarization are plausibly false, with lessons for the empirical study of psychological dynamics

### 2024-04-13 GroupPolarizationStatmod/paper/ms.qmd
expanded plan for next two intro paragraphs to do

*Group polarization* is the name for a group-level process that leads to greater opinion extremity in a group as the group deliberates some topic (e.g., "Do you approve, disapprove, or neither approve nor disapprove of the current administration's policies?"). Extremity only increases, though, assuming that the mean opinion of the group is initially biased towards one extreme or the other [@Brown1986; @Brown2020]. Theoretically, group polarization occurs as groups find consensus, meaning that the variance in group opinion decreases [@DeGroot1974]. Group polarization is important to understand rigorously, therefore, since if groups become increasingly extreme over time, while group opinions are consolidated, this would exacerbate political polarization between political opposition groups, which is already undermines the ability of partisans in democratic governments to work together [@Mason2018UncivilAgreementBook; @Klein2020]. Dozens of studies and meta-analyses over decades have apparently established the empirical reality of group polarization \[@Moscovici1969; @Myers1970; @Isenberg1986; @Sunstein2009; @Sunstein2019\] [@Brown2000, Ch. 5]. However, many or most of the evidence for group polarization relies on experimental designs that use ordinal measurements that are then analyzed as if they were continuous. Unfortunately, this procedure is unfortunately known to induce false positives if it is not properly accounted for [@Liddell2018].

In this paper, we perform a meta-analysis of sixty detections of group polarization from ten published research articles and show that fifty-five of the sixty cannot be reliably identified as group polarization. We show that these 55 detections are equally well-explained by a *simple consensus* model, where the mean group opinion is unchanged by group discussion but opinion variance decreases. We use a novel approach, described herein, that enabled us to identify continuous pre- and post-discussion simple consensus opinion distributions that, when measured with an ordinal scale, result in data that appear to be group-polarized. (How this procedure works in detail: continuous case to identify means, then More broadly, this approach could be used in similar studies that used ordinal measures to determine whether change occurred in distributions of continuous psychological variables. For social psychology and group polarization, this means that an empirical fact must be re-evaluated and new experimental procedures established to rigorously understand whether and when group polarization occurs.

@@

NEXT =\> More about the logic of this study, why showing that these results are plausibly false strictly means they are not rigorously true.

>>>>

*Group polarization* is the name for a group-level process that leads to greater opinion extremity in a group as the group deliberates some topic (e.g., "Do you approve, disapprove, or neither approve nor disapprove of the current administration's policies?"). Extremity only increases, though, assuming that the mean opinion of the group is initially biased towards one extreme or the other [@Brown1986; @Brown2020]. Theoretically, group polarization occurs as groups find consensus, meaning that the variance in group opinion decreases [@DeGroot1974]. Group polarization is important to understand rigorously, therefore, since if groups become increasingly extreme over time, while group opinions are consolidated, this would exacerbate political polarization between political opposition groups, which is already undermines the ability of partisans in democratic governments to work together [@Mason2018UncivilAgreementBook; @Klein2020]. Dozens of studies and meta-analyses over decades have apparently established the empirical reality of group polarization [@Moscovici1969; @Myers1970; @Isenberg1986; @Sunstein2009; @Sunstein2019] [@Brown2000, Ch. 5]. However, many or most of the evidence for group polarization relies on experimental designs that use ordinal measurements that are then analyzed as if they were continuous. Unfortunately, this procedure is unfortunately known to induce false positives if it is not properly accounted for [@Liddell2018].

In this paper, we perform a meta-analysis of sixty detections of group polarization from ten published research articles and show that fifty-five of the sixty cannot be reliably identified as group polarization. We show that these 55 are plausibly false detections because they are equally well-explained by a *simple consensus* model, where the mean group opinion is unchanged by group discussion but opinion variance decreases. We use a novel approach, described herein, that enabled us to identify continuous pre- and post-discussion simple consensus opinion distributions that, when measured with an ordinal scale, result in data that appear to be group-polarized. (How this procedure works in detail: continuous case to identify means, then More broadly, this approach could be used in similar studies that used ordinal measures to determine whether change occurred in distributions of continuous psychological variables. For social psychology and group polarization, this means that an empirical fact must be re-evaluated and new experimental procedures established to rigorously understand whether and when group polarization occurs.

@@

NEXT =\> More about the logic of this study, why showing that these results are plausibly false strictly means they are not rigorously true, overview of methods in somewhat more detail. Maybe need two paragraphs.
